Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-16 paper/sections/acknowledgements.tex
[ paper ] describing simplify.

We thank Joe Leslie-Hurd for his support with its \Metis prover.
We also acknowledge the work carried out by Alejandro G\'omez-Londo\~no~\cite{Gomez-Londono2015} used as the basis code for our \TSTP
parsing module.

Last but not least, we thankfully acknowledge
Andreas Abel and Chalmers University of Goteborg, Sweden for
inviting us to be part of the \Agda Implementors’ Meeting XXV.
We honored to present part of the research in this paper.

>>>>

We thank Joe Leslie-Hurd for his support and comments about \Metis.
We also acknowledge the work carried out by Alejandro
G\'omez-Londo\~no~\cite{Gomez-Londono2015} used as the basis
for our \TSTP parsing module.

Last but not least, we thankfully acknowledge
Andreas Abel and Chalmers University of Goteborg, Sweden for
inviting us to be part of the \Agda Implementors’ Meeting XXV.
We honored to present there part of this research.

### 2017-10-16 paper/sections/conclusions.tex
Paper (#36)
* minor changes.

* [ paper ] section 2 improvements.

* [ paper ] major changes, abstract, conclusions was added.

* [ paper ] minor changes, title and abstract.

* [ paper ] fixing some mistakes.

* [ paper ] fixing error grammars and adding notes.

* [ paper ] section 3 introduction added.

* [ paper ] appendix added.

* added text.

* [ paper ] perliminaries added.

* [ paper ] implementation added, some proofs added.

* [ paper ] fixing some typos and text added.

* [ paper ] fixing some typos and text added.

* [ paper ] complexity measure added.

* [ paper ] minor changes.

In this document, we reconstruct proofs
of propositional logic without equality deliver by \Metis
successfully.

Despite the fact, our logic domain is more straightforward than
first-order or high-order logic where there are many
proof-reconstruction tools (see a list of some of them in the
related work, Section \ref{sec:related-work}), we found some issues
around the reconstruction that show this task is far from trivial.
Depending  on the systems involved in the translation, some
difficulties can arise. We listed some of them in Section
\ref{sec:metis-language-and-proofs} and in Section
\ref{sec:proof-reconstruction}.

For instance, just showing equality between Normal Forms for the
same formula in the output of the \canonicalize inference and a
common implementation for this kind of conversion; we observed that
\Metis performs reordering for each inner expression mainly in the
presence of negative literals. Another aspect to take into account
is the \Metis' treatment for conjunctions and disjunction;
these connectives were treated as right associative and binary
operators in the \TSTP format but that assumptions is not a standard.
We had to update the \TSTP parser module introduced in
\cite{Gomez-Londono2015} to include features like these.

An important matter was how dealing with
the equality between formulas since, in our approach, we only work
with the syntactic treatment of the logic without considering the
semantics.
One of the main reasons for such a restriction relay on our plans
further to support first-order logic, where a semantic treatment is
not viable since it computational complexity.\unsure{Please confirm this.}

Nevertheless, the proof-reconstruction is certainly possible, and
the verification of solutions of \CPL problems can be carried out
successful in \Agda. Our tool is able to translate \Metis
derivations in \TSTP format to deliver \Agda code of the proof,
\change{Better uses \emph{proof-terms}?}
and this proof can be type-checked by \Agda.

The source code of our tool can be extended to support
other \ATPs for \CPL like \name{E}. This work can be carried out
may following the description in Kanso' Ph.D. Thesis~\cite{Kanso2012}
for the inference rules of \name{E} or \name{Z3}.

Trying to justify a proof-object of a theorem prover has a
significant impact, may increase the trustworthiness of the prover
proving its correctness, or contrary, revealing important issues to
improve the system. In both cases, there is a benefit for both
sides, the user of this technology, and their developers.

For instance, while writing this document, we found a soundness bug,
a bug in the printing out of the \TSTP derivations, and a bug in the
parsing for \TPTP problem in \Metis. Fortunately, all these bugs
were fixed quickly by Hurd\footnote{Issues No. 4, No. 2, and the
commit \name{8a3f11e} from the \Metis' official repository
\url{https://github.com/gilith/metis}}, and a new version of \Metis
was released (version 2.3, release 20170810).

% \subsubsection*{Future Work.}
% \label{ssec:future-work}

>>>>

We presented a reliable proof-reconstruction of \Metis solutions
for problems in classical propositional logic without logic.
The reconstruction approach in this study was designed to use
only a syntactic treatment of the logic without considering the
semantics. This aspect represent a difficulty but also an opportunity
to extend this work towards the support for first-order logic.
As a first future work, we see the support of \Metis propositional
fragment with equality.

Summarizing, we provided for each \Metis inference rule a theorem that
emulate the rule deduction. The emulated
version rules exposed in Section~\ref{ssec:emulating-inferences} admit
improvements, we left this for future work, particularly, a better
implementation of \simplify rule.
Some of the improvements are related with computational efficiency.
For instance, we use widely functions to check for syntactical equivalence
between propositions and to reorder propositions to match with another.
We believe both are expensive tasks and can get help in the analysis
step inside the translator tool.

The proof-reconstruction tool presented in this paper
is able to translate \Metis derivations in \TSTP format to \Agda proof-terms.
The development presented in this paper can be extended to support
other \ATPs for \CPL like \name{EProver} or \name{Z3}.
This future work can be carried out by following the \name{EProver} inference
rules described on Kanso' Ph.D. Thesis~\cite{Kanso2012}.

One of the main contribution of this study is
increasing the trustworthiness of these automatic provers,
in our case, \Metis. Justify a proof-object of a theorem prover has a real
significant impact for these automatic tools.
The reverse engineering task can reveal important issues or bugs
in many parts of these system (e.g., preprocessing, reasoning, or
deduction modules). While doing this research, we had the opportunity
to find some of those kind of bugs\footnote{Issues No.~2, No.~4. and
commit \name{8a3f11e} in the \Metis' official repository
\url{https://github.com/gilith/metis}}. Fortunately, all these bugs
were fixed quickly for release 20170810 by Joe Hurd.

### 2017-10-16 paper/sections/proof-reconstruction.tex
[ paper ] describing simplify.

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the same support of type theory and inductive types is valid (see a description of \Agda for its main features in
Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}, the process begin with
a \TPTP file that encodes a problem in \CPL. We use this file as the
input of the \Metis prover and if the problem is a theorem, \Metis can reply with a derivation in \TSTP format of the proof.

@@

\subsubsection{A Practical Example.}
Let us consider the following example to exhibit the technique
mentioned above for an inner function present in the emulated version
of the \strip inference.

@@

\subsubsection{Strip a Goal.}
\label{sssec:strip-a-goal}

@@

\subsubsection{Negate.}

@@

The \resolve rule incorporates the resolution theorem showed in
Fig.~\ref{fig:metis-inferences} for the \TSTP derivations.
This rule posses three arguments: a positive literal $l$ to apply the resolution theorem, and two derivations,
$\mathcal{D}₁$ and $\mathcal{D}₂$.
The positive literal $l$ must occur in
the formula from the first derivation $\mathcal{D}₁$, and the
negative literal must occur in the formula from the second derivation $\mathcal{D}₂$ (see the pattern in the \emph{resolve} rule).
The following excerpt is an \Metis \TSTP example for the \resolve rule.

\begin{verbatim}
cnf(refute_0_0, plain, p | q,
  inference(canonicalize, [], [normalize_0_0])).
cnf(refute_0_1, plain, ~ p,
  inference(canonicalize, [], [normalize_0_4])).
cnf(refute_0_2, plain, q,
  inference(resolve, [$cnf(p)], [refute_0_0, refute_0_1])).
\end{verbatim}

To emulate this rule, we perform rearrangements inside the formulas
involved to match with the pattern expected by the resolution theorem
described above.

\begin{definition}[rsol]
\begin{align*}
      \begin{split}
        &rsol : \Prop \to \Prop\\
        &rsol(φ) =
        \begin{cases}
          φ₂, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁, φ₂ ≡ φ₄\\
          φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁\\
          φ, &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{definition}

@@

\begin{definition}[resolve]
 \begin{equation*}
    resolve(φ₁, φ₂, l, ψ) =
      rsol(reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨} (φ₂, ¬ l ∨ ψ)).
  \end{equation*}
\end{definition}

@@

@@

The \canonicalize rule is an overloaded inference rule that performs
mainly normalization of the proposition input by using theorems
described in the section above. This rule  can convert a FOF formula in
clausal form to a CNF clause. \Metis can introduce axioms, and definitions
by using this rule. Since the nature of this rule consists of dealing with
CNF clauses, our strategy to justify its reasoning mainly consists of
testing for equality with normalized negative form of the formula, or a
reordering of the conjunctive normal form of the source formula that
matches the conjunctive normal form of the target formula.

\begin{definition}[canonicalize]
\begin{align*}
    \begin{split}
      &canonicalize : \Prop \to \Prop \to \Prop\\
      &canonicalize(φ, ψ) =
      \begin{cases}
        ψ, &\text{if }ψ ≡ φ\\
        ψ, &\text{if }ψ ≡ form(nnf(φ))\\
        ψ, &\text{if }cnf(ψ) ≡ reorder_{∧∨}(cnf(φ), cnf(ψ))\\
        φ, &\text{otherwise.}\\
      \end{cases}
    \end{split}
\end{align*}
\end{definition}

@@

\begin{remark}
The \clausify rule is always preceded by the application of the
\canonicalize rule. Both rules perform Clausification that introduces
into the problem axioms or definitions in the domain of the solution.
The process is mainly described on paper by \citeauthor{Sutcliffe1996}~\cite{Sutcliffe1996}.
\end{remark}

@@

% \begin{lemma}[lem-canon]
%   \label{lem:lem-canon}
%   \begin{align*}
%   \begin{split}
%   &canon(φ) =\\
%   &\hspace{3mm}
%   \begin{cases}
%       canon(φⱼ), &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊤,\\
%                  &\text{for same } i,j = 1,2, i≠j\\
%      ⊥, &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊥, i = 1,2\\
%      canon(φ₁) ∧ canon(φ₂), &\text{if } φ ≡ φ₁ ∧ φ₂\\
%      ⊤, &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊤, i = 1,2\\
%      canon(φⱼ), &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊥,\\
%                 &\text{for same } i,j = 1,2, i≠j\\
%      canon(φ₁) ∨ canon(φ₂), &\text{if } φ ≡ φ₁ ∨ φ₂\\
%      φ, &\text{otherwise.}
%     \end{cases}
%   \end{split}
%   \end{align*}
% \end{lemma}

% \begin{theorem}[thm-canonicalize] % (fold)
%   \label{thm:thm-canonicalize}
%   $Γ ⊢ φ$ then $Γ ⊢ canonicalize(φ)$, where,
%   \begin{equation*}
%     canonicalize =
%      canon ∘ ndisj_{∧} ∘ ndisj_{∧∨} ∘ redun ∘ rconj ∘ cnf.
%   \end{equation*}
% \end{theorem}

% The \canonicalize rule acts in \Metis' \TSTP derivations in three main cases. First, introducing the
% negation of a subgoal. Second, to simplify an intermediate derivation.
% Third, introducing an axiom or a definition of the.
% To justify the two first cases, we use the Theorem~\ref{thm:thm-canonicalize}. For the last case, we use the Lemma~\ref{lem:lem-nnf}.

% -------------------------------------------------------------------
\subsubsection{Simplify.}
\label{sssec:simplify}

The \simplify rule is an inference that performs simplification of
definitions and tautologies. This rule
transverses a list of formulas by applying different theorems (e.g.,
\emph{modus pones}, \emph{modus tollens}, or \emph{disjunctive
syllogism}) including those theorems for removing redundancies and definitions  from the description of the \canonicalize rule,
Fig. \ref{fig:conjunctive-disjunctive-simpl}. This rules also use the resolution as we show in the description of the \resolve rule~\ref{thm:thm-resolve}. Let see the following excerpt from a \Metis \TSTP derivation where \simplify appears.

\begin{verbatim}
fof(normalize_1_0, plain, (~ p & ~ r & q),
    inference(canonicalize, [], [negate_1_0])).
fof(normalize_1_1, plain, ((p | q) & (p | r)),
    inference(canonicalize, [], [a1])).
fof(normalize_1_2, plain, (p | r),
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_3, plain, ($false),
    inference(simplify, [], [normalize_1_0, normalize_1_2])).
\end{verbatim}

>>>>

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the same support of type theory and inductive types is valid.
%  (see a description of \Agda for its main features in
% Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}, the process begin with
a \TPTP file that encodes a problem in \CPL. We use this file as the
input of the \Metis prover and if the problem is a theorem, \Metis can reply with a derivation in \TSTP format of the proof.

@@

\subsubsection*{A Practical Example.}
Let us consider the following example to exhibit the technique
mentioned above for an inner function present in the emulated version
of the \strip inference.

@@

\subsubsection{Stripping a Goal.}
\label{sssec:strip-a-goal}

@@

\subsubsection{Negating Subgoals.}

@@

The \resolve rule incorporates the resolution theorem showed in
Fig.~\ref{fig:metis-inferences} for the \TSTP derivations.
This rule posses three arguments: a positive literal $l$ to apply the
resolution theorem, and two derivations.
The positive literal $l$ must occur in
the formula from the first derivation and the
negative literal must occur in the formula from the second derivation,
see the pattern of \emph{resolve} rule in Fig.~\ref{fig:metis-inferences}.

% \begin{figure}
\label{fig:resolve-tstp-example}
\begin{verbatim}
cnf(refute_0_4, plain, (~ r | p | q),
    inference(canonicalize, [], [normalize_0_5])).
cnf(refute_0_5, plain, (p | q | r),
    inference(canonicalize, [], [normalize_0_6])).
cnf(refute_0_6, plain, (p | q),
    inference(resolve, [$cnf(r)], [refute_0_5, refute_0_4])).
\end{verbatim}
% \caption{Excerpt of a \Metis \TSTP example for the \resolve rule.}
% \end{figure}

To emulate this rule, we perform rearrangements inside the formulas
involved to match with the pattern expected by the resolution theorem
described above. We begin by defining a convenient version of
the binary resolution in Def.~\ref{def:rsol}.

\begin{definition}[rsol]
\label{def:rsol}
\begin{align*}
      \begin{split}
        &rsol : \Prop \to \Prop\\
        &rsol(φ) =
        \begin{cases}
          φ₂, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁, φ₂ ≡ φ₄\\
          φ₂ ∨ φ₄, &\text{ if }φ ≡ (φ₁ ∨ φ₂) ∨ (φ₃ ∨ φ₄), φ₃ ≡ ¬ φ₁\\
          φ, &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{definition}

@@

In Def.~\ref{def:resolve}, the emulated rule for $resolve$
has four arguments. The first three were described
above. The last one, ψ, represents the expected conclusion in the \Metis
derivation.

\begin{definition}[resolve]
\label{def:resolve}
 \begin{equation*}
    resolve(φ₁, φ₂, l, ψ) =
      rsol(reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨} (φ₂, ¬ l ∨ ψ)).
  \end{equation*}
\end{definition}

@@

\begin{figure}
\label{fig:resolve-example}
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ ¬ r ∨ p ∨ q$}
  \AxiomC{$Γ ⊢ p ∨ q ∨ r $}
  \RightLabel{thm-resolve}
  \BinaryInfC{$Γ ⊢ resolve(¬ r ∨ p ∨ q, p ∨ q ∨ r, r, p ∨ q)$}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \end{bprooftree}
\end{equation*}
\caption{Example of a resolution in \Metis using
Thm.~\ref{thm:thm-resolve}.}
%Application of Thm.~\ref{thm:thm-resolve} for the \TSTP excerpt Fig.~\ref{fig:resolve-tstp-example}.}
\end{figure}

@@

The \canonicalize rule is an overloaded inference rule that performs
mainly normalization of the proposition input by using theorems
described in the section above. This rule  can convert a FOF formula in
clausal form to a CNF clause. \Metis can introduce axioms, and definitions
by using this rule. Since the nature of this rule consists of dealing with
CNF clauses, our strategy to justify its reasoning mainly consists of
testing for equality with normalized negative form of the formula, or a
reordering of the conjunctive normal form of the source formula to match
the conjunctive normal form of the target formula.

\begin{definition}[canonicalize]
\label{def:canonicalize}
\begin{align*}
    \begin{split}
      &canonicalize : \Prop \to \Prop \to \Prop\\
      &canonicalize(φ, ψ) =
      \begin{cases}
        ψ, &\text{if }ψ ≡ φ\\
        ψ, &\text{if }ψ ≡ form(nnf(φ))\\
        ψ, &\text{if }cnf(ψ) ≡ reorder_{∧∨}(cnf(φ), cnf(ψ))\\
        φ, &\text{otherwise.}\\
      \end{cases}
    \end{split}
\end{align*}
\end{definition}

@@

\begin{remark}
The \clausify rule is often preceded by the \canonicalize rule.
Both rules perform \emph{Clausification} that introduces
into the problem axioms or definitions in the domain of the solution.
The Clausification algorithms is mainly described on paper by
\citeauthor{Sutcliffe1996}~\cite{Sutcliffe1996}. We left for future work
investigate the consequences of removing $clausify$ inference by
strengthen $canonicalize$ rule in Def.~\ref{def:canonicalize}
\end{remark}

@@

% -------------------------------------------------------------------
\subsubsection{Simplification.}
\label{sssec:simplify}

The \simplify rule is an inference that performs simplification of
definitions and tautologies. This rule
transverses a list of inner derivations by applying different theorems
(e.g., \emph{modus pones}, \emph{modus tollens}, or \emph{disjunctive
syllogism}) to get a contradiction or just simplify the formulas.
To emulate this rule, we have included theorems
from \canonicalize rule Section~\ref{sssec:canonicalize} and
\resolve rule Section~\ref{sssec:resolve}.

\begin{remark}
Besides the fact, $\List\ \Prop \to \Prop$ is the type that coincides
with the reading of a \TSTP derivation, we choose a different option.
In the translation from \TSTP to \Agda, we build a conjunction
from the list of derivations by introducing binary conjunction
with $∧$-intro rule. That avoids us to define a new sequent type
to support \List \Prop type in the conclusion side.
\end{remark}

\begin{figure}
\begin{verbatim}
fof(normalize_1_0, plain, (~ p & ~ r & q),
    inference(canonicalize, [], [negate_1_0])).
fof(normalize_1_1, plain, ((p | q) & (p | r)),
    inference(canonicalize, [], [a1])).
fof(normalize_1_2, plain, (p | r),
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_3, plain, ($false),
    inference(simplify, [], [normalize_1_0, normalize_1_2])).
\end{verbatim}
\caption{Example of \simplify rule in a \TSTP derivation of \Metis.}
\end{figure}

\begin{align}
    \begin{split}
      &\textit{simpl} : \Prop \to \Prop \to \Prop \to \Prop\\
      &\textit{simpl}(φ₁, φ₂ ψ) =
      \begin{cases}
        ψ, &\text{if }φᵢ ≡ ψ,\text{ for some }i = 1, 2\\
        ψ, &\text{if }φ₁ ≡ φ₂ ⇒ ψ\\
      \end{cases}
    \end{split}
\end{align}

\begin{theorem}[thm-simplify] % (fold)
  \label{thm:thm-simplify}
$\{ Γ ⊢ φᵢ \}$ for $i = 1, \cdots, n$ then
$Γ ⊢ simplify (\bigwedge_{i=1}^n φ)$.
\end{theorem}

### 2017-10-16 paper/sections/related-work.tex
[ paper ] describing simplify.

% Kanso and Setzer~\cite{kanso2016light} integrate \name{Z3} in \Agda
% but they also integrate the propositional fragment of the \name{E} prover in
% \cite{Kanso2012}. They cataloged these two integration as
% \emph{Oracle and Reflection} and \emph{Oracle and Justification}, respectively.
% Their integration with the \name{E} prover is the most related work with our
% approach found at the moment while we write this document. Besides the use
% of the same \ITP, crucial differences easily arise like decidability of
% propositional logic using semantics that we do not take into account for the
% reason we describe later on in Section~\ref{sec:conclusions}.

\end{document}

>>>>

Kanso and Setzer~\cite{kanso2016light} integrate the propositional fragment of \name{EProver} in \Agda. In this work, the authors translate automatically propositional logic problems written in \Agda to \TPTP problems. By calling an external wrapper the ATP, this tool gets a \TSTP derivation. The tool parses the \name{EProver} derivation and outputs
\Agda code. After that, we say their proof-reconstruction is almost
similar as the approach presented in Section~\ref{ssec:workflow}.
Nevertheless, Kanso and Setzer aim to justify some of the inference rules
by using propositional semantics instead of just syntactic derivation
steps.
\end{document}
